---
title: "Data Wrangling"
description: "Data wrangling projects using Tidyverse, dplyr, tidyr, and ggplot2."
---

This section documents data wrangling coursework including data reshaping, pivot operations, missing value handling, and multi-dataset analysis using the Tidyverse ecosystem.

## Air Quality Data

```{r}
#| echo: true
#| warning: false
#| message: false

library(tidyverse)

air.tib <- as_tibble(airquality)

# Filter and select
air.tib %>%
  filter(Temp == 97) %>%
  select(Ozone)
```

## NYC Flights: Worst Arrival Delays

```{r}
#| echo: true
#| warning: false
#| message: false

library(nycflights13)

worst10_arr_delays <- flights %>%
  filter(!is.na(arr_delay)) %>%
  group_by(day) %>%
  slice_max(arr_delay, n = 10, with_ties = FALSE) %>%
  ungroup() %>%
  group_by(month) %>%
  summarise(avg_delay = mean(arr_delay, na.rm = TRUE))

ggplot(worst10_arr_delays, aes(x = factor(month), y = avg_delay)) +
  geom_col(fill = "cyan") +
  labs(
    title = "Average of Top 10 Worst Arrival Delays per Day by Month",
    x = "Month",
    y = "Average Arrival Delay (Minutes)"
  ) +
  theme_minimal()
```

The worst arrival delays happen in the beginning of the year, with the worst occurring in February. This suggests the company can allocate more budget to the first two quarters.

## Religion & Income: Pivot Longer

```{r}
#| echo: true

relig_income_long <- relig_income %>%
  pivot_longer(
    cols = -religion,
    names_to = "income",
    values_to = "count"
  ) %>%
  filter(income != "Don't know/refused")

ggplot(relig_income_long, aes(x = income, y = count, fill = religion)) +
  geom_col(position = "dodge") +
  labs(
    title = "Income Distribution by Religion",
    x = "Income Bracket",
    y = "Number of Respondents"
  ) +
  theme_minimal()
```

## US Rent vs Income: Pivot Wider

```{r}
#| echo: true

us_rent_income %>%
  pivot_wider(
    names_from = variable,
    values_from = c(estimate, moe)
  ) %>%
  ggplot(aes(x = estimate_income, y = estimate_rent)) +
  geom_point(color = "blue", size = 3) +
  geom_smooth(method = "lm", se = FALSE, color = "orange") +
  labs(
    title = "Income vs Rent Across States",
    x = "Income (Median)",
    y = "Rent (Median)"
  ) +
  theme_minimal()
```

## Missing Value Handling

```{r}
#| echo: true

air.tib %>%
  summarise(
    missing = sum(is.na(Solar.R)),
    percent_missing = mean(is.na(Solar.R)) * 100
  )

Solar_Mean <- mean(air.tib$Solar.R, na.rm = TRUE)
Solar_Mean_int <- round(Solar_Mean)

air.tib <- air.tib %>%
  mutate(Solar.R = replace_na(Solar.R, Solar_Mean_int))

sum(is.na(air.tib$Solar.R))
```

## Temperature Analysis by Month

```{r}
#| echo: true

air.tib %>%
  count(Month, wt = Temp) %>%
  mutate(
    Month = recode(Month,
                   `5` = "May",
                   `6` = "June",
                   `7` = "July",
                   `8` = "August",
                   `9` = "September"),
    Month = factor(Month)
  ) %>%
  arrange(desc(n)) %>%
  ggplot(aes(x = n, y = reorder(Month, n))) +
  geom_col(fill = "cyan") +
  labs(
    title = "Total Monthly Temperature",
    x = "Total Temperature",
    y = "Month"
  ) +
  theme_classic()
```

## Ozone & Weather Variable Relationships

```{r}
#| echo: true

air.tib %>%
  mutate(temp_wind = Temp / Wind) %>%
  select(Ozone, Temp, Wind, temp_wind) %>%
  pivot_longer(cols = c(Temp, Wind, temp_wind),
               names_to = "variable",
               values_to = "value") %>%
  ggplot(aes(x = value, y = Ozone)) +
  geom_point(color = "steelblue") +
  facet_wrap(~variable, scales = "free_x") +
  labs(
    title = "Relationships Between Ozone and Weather Variables",
    x = "Variable Value",
    y = "Ozone Level"
  ) +
  theme_minimal()
```

There is a stronger relationship between ozone level and temperature, and ozone level and wind, than between ozone level and temp_wind. There is also a negative relationship between wind and ozone level.

## Key Wrangling Functions

| Function | Purpose |
|---|---|
| `pivot_longer()` | Reshape wide data to long format |
| `pivot_wider()` | Reshape long data to wide format |
| `filter()` | Subset rows based on conditions |
| `mutate()` | Create or modify columns |
| `group_by() + summarise()` | Aggregate data by groups |
| `replace_na()` | Handle missing values |
| `across()` | Apply calculations across multiple columns |
| `distinct()` | Remove duplicate rows |
| `inner_join()` | Merge datasets on matching keys |
