---
title: "R Programming"
description: "R programming projects covering functions, purrr, forcats, lubridate, and functional programming."
---

This section documents R programming coursework covering data types, custom functions, functional programming with purrr, factor handling with forcats, and date-time operations with lubridate.

## Working with Vectors and Data Frames

```{r}
#| echo: true
#| warning: false
#| message: false

library(tidyverse)

# Create vectors
stock_prices <- c(42.5, 43.0, 44.1, 2.7)
stock_company <- c("Microsoft", "Apple", "Facebook", "Tesla")
stock_ticker <- c("MSFT", "AAPL", "FB", "TSLA")
make_cars <- c(FALSE, FALSE, FALSE, TRUE)

# Update and name
stock_prices <- stock_prices + 5
names(stock_prices) <- stock_company

# Build data frame
stock_info.df <- data.frame(
  stock_prices,
  stock_company,
  stock_ticker,
  make_cars,
  stringsAsFactors = FALSE
)

# Correct Tesla
stock_info.df[stock_info.df$stock_company == "Tesla", "stock_prices"] <- 47.7

# Add new row
extra <- data.frame(
  stock_prices = 50.4,
  stock_company = "General Motors",
  stock_ticker = "GM",
  make_cars = TRUE
)
stock_info.df <- rbind(stock_info.df, extra)
stock_info.df$open_price <- c(47.3, 47.8, 49.5, 47.9, 55.3)

# Percentage change
stock_info.df$change_percent <-
  ((stock_info.df$stock_prices - stock_info.df$open_price) /
     stock_info.df$open_price) * 100

stock_info.df
```

## Functional Programming with purrr

The purrr package simplifies iteration by replacing traditional loops with cleaner functions like `map()`. Instead of calculating the mean of each column separately, `map_dbl()` computes all means in a single step.

```{r}
#| echo: true

library(purrr)

cols <- c("mpg", "hp", "wt")
map_dbl(mtcars[cols], mean)
```

## Factor Handling with forcats

```{r}
#| echo: true

library(gapminder)
library(forcats)

# Population by continent in 2007
pop_2007 <- gapminder |>
  filter(year == 2007) |>
  summarise(total_pop = sum(pop), .by = continent)

pop_2007 |>
  arrange(desc(total_pop))

# Reorder factor levels
gapminder |>
  filter(year == 2007) |>
  summarise(total_pop = sum(pop), .by = continent) |>
  ggplot(aes(fct_reorder(continent, total_pop), total_pop)) +
  geom_col() +
  coord_flip() +
  labs(title = "Population by Continent (2007)",
       x = "Continent",
       y = "Population")
```

## Top & Bottom Countries: Life Expectancy

```{r}
#| echo: true

library(patchwork)

asia_2007 <- gapminder |>
  filter(year == 2007, continent == "Asia")

# Top 5
top_plot <- asia_2007 |>
  slice_max(lifeExp, n = 5) |>
  ggplot(aes(fct_reorder(country, lifeExp), lifeExp)) +
  geom_col() +
  coord_flip() +
  labs(title = "Top 5 Countries")

# Bottom 5
bottom_plot <- asia_2007 |>
  slice_min(lifeExp, n = 5) |>
  ggplot(aes(fct_reorder(country, lifeExp), lifeExp)) +
  geom_col() +
  coord_flip() +
  labs(title = "Bottom 5 Countries")

top_plot + bottom_plot
```

## Date-Time Operations with lubridate

```{r}
#| echo: true

library(nycflights13)
library(lubridate)

flights_lub <- flights |>
  filter(!is.na(dep_time), !is.na(arr_time), !is.na(air_time)) |>
  mutate(
    dep_time = make_datetime(year, month, day, dep_time %/% 100, dep_time %% 100),
    arr_time = make_datetime(year, month, day, arr_time %/% 100, arr_time %% 100)
  ) |>
  select(origin, dest, air_time, starts_with("dep"), starts_with("arr"))

# Daily flight distribution
flights_lub |>
  count(date = as.Date(dep_time)) |>
  ggplot(aes(date, n)) +
  geom_line() +
  labs(title = "Flights per Day (2013)")
```

## String Operations

```{r}
#| echo: true

library(stringr)

# Extract first words from sentences
first_words <- str_trim(word(sentences, 1))
head(first_words, 20)

# Count "The" as first word
count_the <- sum(first_words == "The")
prop_the <- mean(first_words == "The")

cat("Count:", count_the, "\n")
cat("Proportion:", round(prop_the, 3), "\n")
```

## Key Programming Concepts

::: {.project-grid}
::: {.project-card}
### Custom Functions

Define reusable functions to reduce repetition and make code more readable. Whenever code is repeated more than twice, convert it into a function.

<span class="tag">DRY Principle</span>
:::

::: {.project-card}
### Functional Iteration

Use `map()`, `map_dbl()`, and `map_dfr()` from purrr instead of traditional for loops for cleaner, more expressive code.

<span class="tag">purrr</span>
:::

::: {.project-card}
### Factor Reordering

Use `fct_reorder()` from forcats to arrange categorical variables by a numeric summary for more meaningful visualizations.

<span class="tag">forcats</span>
:::

::: {.project-card}
### Anonymous Functions

Define functions inline without naming them. This makes code more concise for one-time operations inside pipes.

<span class="tag">Functional</span>
:::
:::
